---
title: "Pixar Films"
description: |
  Analysis of Pixar film data from TidyTuesday
author: Anjali Suva
date: September 23, 2025
format:
  html:
    code-fold: true
    code-summary: "Show the code"
execute: 
  warning: false
  message: false
---

For this first mini-project, I wanted to examine data from a TidyTuesday data set on Pixar films. The data set provides information on the name of the film, order of release, the date the film premiered, run time, and the film's rating based on the Motion Picture Association (MPA)'s rating system. More information on the data set can be found from the sources, linked below.

For my initial exploration, I decided to visualize the distribution of run times for Pixar films. Since I'm only interested in one variable, I chose to create a boxplot.

```{r}
#| fig-alt: "Box plot of Pixar films by run time, where median run time is around 100 minutes with two outliers."

#load necessary packages for my data exploration + custom ggplot theme
library(tidytuesdayR)
library(ggplot2)
library(readr)
source("R/theme_asuva.R")

#loading the data set
pixar_films <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-03-11/pixar_films.csv")

#creating a box plot for film length 
ggplot(pixar_films, aes(x = run_time))+
  geom_boxplot(color="#5e4a4a")+
  labs(x ="Run time (minutes)",title ="Length of Pixar Films") +
  theme_asuva()

```

According to the boxplot, the median run time for Pixar films was around 100 minutes. Two films in the set are outliers, causing the graph to skew right.

I wanted to know what the values were for the median and the first and third quartile, so I wrangled the data to get the summary statistics. The tibble below shows what I found:

```{r}

library(dplyr)

pixar_films |> 
  summarize(average_runtime = median(run_time, na.rm = TRUE), Q1 = quantile(run_time, 0.25, na.rm = TRUE), Q3 = quantile(run_time, 0.75, na.rm = TRUE))

```

I also wanted to see what films were the two potential outliers in the box plot. To do this, I created code for a data frame using the mutate function that would alert me to any outliers if the film's run time was over 120 minutes, a metric I obtained from the boxplot above. The tibble is shown below. One of the outliers was *Luca*, with a run time of 151 minutes. The other film was unknown from the data set, but after some research done using the info in the release_date column, I found out that the film was *Elemental*.

```{r}
pixar_films |>
  select(film, run_time, release_date) |>
  mutate(potential_outlier = ifelse(run_time > 120, "Yes", "No")) |>
  arrange(desc(potential_outlier))
```

[Credits:]{.underline}

**Data:** TidyTuesday Pixar Films, <https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-03-11/readme.md>

**Original Source:** pixarfilms R package by Eric Leung, drawing data from Wikipedia and OMDb, <https://github.com/erictleung/pixarfilms>
